---
title: 'Bellabeat Case Study: FitBit Usage Trend Analysis'
author: "Masaki Nakanishi"
date: "2022-09-01"
output:
  pdf_document: default
  html_document: default
---

## Introduction

This project aims to investigate how people use existing smart devices and to give a high-level recommendation for how the findings can inform Bellabeat marketing strategy. 

Bellabeat is a high-tech company that manufactures health-focused smart devices including bracelet, necklace, clip, watch, and water bottle. The smart bracelet, necklace, clip, and watch can track users' activity, sleep and stress. The smart water bottle can track users' daily water intake. All the smart devices can be connected to Bellabeat app so that users can review their comprehensive health-related habits and information. 

In this project, we, Bellabeat marketing analytics team, analyzed a publicly-available dataset to gain an insight into how people who alreay own smart devices use them, and revealed potential marketing opportunities for Bellabeat's growth. 

## Questions

To identify some trends in smart device usage, we will aim to answer the following questions:

* Do people use their smart devices for tracking specific information (e.g., activity, sleep, and stress)? For example, do they use their devices for tracking activity rather than stress and sleep, or vice versa? 

* Assuming that most people use their devices for tracking their activity, during what kinds of activities (e.g., light, moderate, and intensive activities) do they use them the most? 

To address the questions, we will conduct a series of studies. 

## Study 1

### Hypothesis

* People use their smart devices for tracking activity or stress rather than sleep.

    Since sleep quality can be tracked by different devices such as a smartphone, people who bought a wearable smart device might be interested in the information that can only be tracked by wearable devices.
    
### Materials and Methods

We used FitBit Fitness Tracker Data (https://www.kaggle.com/datasets/arashnic/fitbit) to test the hypothesis. FitBit is also a company that manufactures smart devices. Their smart devices can also track users' activity and sleep. The FitBit Fitness Tracker Data contains personal tracker data including physical activity, heart rate, and sleep monitoring obtained from ## participants recorded between 04-12-2016 - 05-12-2016. 

Using the data, we compared the participants' usage of the FitBit for monitoring their physical activity and sleep. The metrics we used were:
* The number of days each participant used to track his/her physical activity
* The number of days each participant used to track his/her sleep

We submitted the two metrics into a paired t-test to test a null-hypothesis: there is no difference in the mean numbers of days used to track phsyical activity and sleep. 

### Analysis

#### Data cleaning in SQL
We first cleaned the data using SQL. We removed invalided observations where all the columns are 0 (indicating he/she did not use his/her device on that day) and computed the metrics (the numbers of days each participant used to track his/her physical activity and sleep) using the raw daily usage information in Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv and Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv using the following queries.

SELECT
  ID, COUNT(ActivityDate) AS NumOfActiveDays
FROM
  `bellabeat-case-study-361204.FitBit_Fitness_Tracker_Data.Daily_Activity`
WHERE
  TotalDistance != 0
GROUP BY
  ID

SELECT
  ID, COUNT(SleepDay) AS NumOfSleepDays
FROM
  `bellabeat-case-study-361204.FitBit_Fitness_Tracker_Data.Daily_Sleep`
WHERE
  TotalSleepRecords != 0
GROUP BY
  ID

We then saved the tables as numOfDaysForActivityTrack.csv and numOfDaysForSleepTrack.csv.

#### Preparation

A package that was necessary for the following analysis was installed as follows:

```{r}
install.packages("tidyverse", repos="http://cran.us.r-project.org")
library(tidyverse)
```

Two data tables were imported and merged together. In addtion, the data cells that contained NA were replaced by 0, and a new column that contained the participant ID in a character type was added.

```{r}
activity_data = read_csv("Fitabase_Data_4_12_16-5_12_16/numOfDaysForActivityTrack.csv")
sleep_data = read_csv("Fitabase_Data_4_12_16-5_12_16/numOfDaysForSleepTrack.csv")
data_merged = merge(x = activity_data, y = sleep_data, by = "ID", all = TRUE)
data_merged = mutate(data_merged, char_ID = as.character(data_merged$ID))
data_merged[is.na(data_merged)] <- 0
head(data_merged)
```

For the following visualization, the merged data were reformated to a long table with three columns including participant ID, purpose of using the device (i.e., Activity or Sleep), and the number of days they used the device for a purpose. 

```{r}
long_data = data.frame(Participant =  c(data_merged$char_ID, data_merged$char_ID), Purpose = rep(c('Activity', 'Sleep'), each = length(data_merged$ID)), NumOfDays = c(data_merged$NumOfActiveDays, data_merged$NumOfSleepDays))
head(long_data)
```

We also computed the averaged number of days for each purpose across participants and its standard errors.

```{r}
sum_data = data.frame(Purpose = c("Activity","Sleep"), Average = c(mean(data_merged$NumOfActiveDays), mean(data_merged$NumOfSleepDays)), SE = c(sd(data_merged$NumOfActiveDays)/sqrt(length(data_merged$ID)), sd(data_merged$NumOfSleepDays)/sqrt(length(data_merged$ID))))
sum_data
```
#### Visualization

The following bar plot indicates the number of days each participant used the device for tracking activity and sleep.

```{r}
ggplot(long_data, aes(y = Participant, x = NumOfDays, fill=Purpose)) + 
  geom_bar(position = 'dodge', stat = "identity", width = 0.8) +
  labs(title="The numbers of days used for tracking for each individual", y = "Participants", x = "The number of days") 
```

The following bar plot indicates their average across participants. The error bars indicate the standard error.

```{r}
ggplot(sum_data, aes(x=Purpose, y=Average)) + 
  geom_bar(stat="identity", position=position_dodge()) + 
  geom_errorbar(aes(ymin=Average-SE, ymax=Average+SE), width=.1, position=position_dodge(.9)) +
  labs(title="Averaged number of days used for tracking across participants", x="Purpose", y="Averaged number of days")
```
As the bar chart indicated, people tended to use their devices for tracking activity rather than sleep. A paired t-test revealed that there was a statistically signiticant difference between the activity and sleep (p < 0.001).

```{r}
t.test(data_merged$NumOfActiveDays, data_merged$NumOfSleepDays, paired=TRUE)
```


```{r}
#r = cor(data_merged$NumOfActiveDays, data_merged$NumOfSleepDays)
#ggplot(data_merged, mapping=aes(x=NumOfActiveDays, y=NumOfSleepDays)) + geom_point() +
#  geom_smooth(method="lm") + 
#  labs(title="Association between the numbers of days used for tracking activity and sleep", x="The number of days #(Activity)", y = "The number of days (Sleep)") +
#  annotate(geom = "text", x = 4, y = 31, label=sprintf("R = %.2f", r))
```

### Summary

The results showed that people tended to use FitBit device for tracking their activity rather than their sleep, which supported our hypothesis. Note that the participants tracked their activity on 26.12 $\pm$ 1.19 days out of 31 days on average. On the other hand, they tracked their sleep only on 12.52 $\pm$ 2.17 days on average. This indicates that people are much more intersted in monitoring their activity than their sleep. 


## Study 2

We found that most of the FitBit users use their devices for tracking their activities. Now we are interested in investigating what kind of activities (e.g., light, moderate, and intensive) they perform while using the devices.

### Hypothesis

* People use their smart devices for tracking light activity rather than moderate and intensive activities. 

Since light activity such as walking is a part of daily activity, people use their smart devices mostly while performing light activity.

### Materials and Methods

Again, we used FitBit Fitness Tracker Data (https://www.kaggle.com/datasets/arashnic/fitbit) to test the hypothesis.

Using the data, we compared the duration of using their FitBit for tracking different types of activity. The metrics we used were:
* Total lightly, fairly, and very active duration in minutes.
* Ratio of duration of each types of activities to the total active duration.

We submitted the two metrics into a one-way repeated measures analysis of variance (ANOVA) to test a null-hypothesis: there is no difference in the mean active durations among three types of activities. 

### Analysis

#### Data cleaning in SQL

We removed invalided observations where all the columns are 0 (indicating he/she did not use his/her device on that day) and computed the metrics (total lightly, fairly, and very active duration in minutes, and Ratio of duration of each types of activities to the total active duration) using the raw daily usage information in Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv using the following queries.

WITH data_converted AS (
  SELECT
    ID, 
    sum(LightlyActiveMinutes) AS LightlyActiveMinutesTotal,
    sum(FairlyActiveMinutes) AS FairlyActiveMinutesTotal,
    sum(VeryActiveMinutes) AS VeryActiveMinutesTotal,
    (sum(LightlyActiveMinutes)+sum(FairlyActiveMinutes)+sum(VeryActiveMinutes)) AS ActiveMinutesTotal
  FROM
    `bellabeat-case-study-361204.FitBit_Fitness_Tracker_Data.Daily_Activity`
  WHERE
    TotalDistance != 0
  GROUP BY ID
  ORDER BY ID
)

SELECT
  ID, 
  LightlyActiveMinutesTotal,
  FairlyActiveMinutesTotal,
  VeryActiveMinutesTotal,
  ActiveMinutesTotal,
  LightlyActiveMinutesTotal/ActiveMinutesTotal*100 AS LightlyActivePerc,
  FairlyActiveMinutesTotal/ActiveMinutesTotal*100 AS FairlyActivePerc,
  VeryActiveMinutesTotal/ActiveMinutesTotal*100 AS VeryActivePerc
FROM data_converted

We then saved the table as activityLogPercentage.csv for the following analysis in R. In R, the data table looks like as follows:

```{r}
activity_merged = read_csv("Fitabase_Data_4_12_16-5_12_16/activityLogPercentage.csv")
head(activity_merged)
```

We converted the data table to a long format from a wide format and added a new column that contained the participant ID in a character type was added as follows:

```{r}
activity_wide = summarize(activity_merged, ID, LightlyActivePerc, FairlyActivePerc, VeryActivePerc)
activity_long = gather(activity_wide, Activity, Percentage, -ID)
activity_long = mutate(activity_long, char_ID = as.character(activity_long$ID))
head(activity_long)
```

The followin chart indicates how long each participants have used their devices for tracking light, moderate, and intensive activity in percentage. As shown in the chart, all the participants mostly used their devices for monitoring light activity.

```{r}
ggplot(activity_long, aes(y = char_ID, x = Percentage, fill=Activity)) + 
  geom_col(position="fill", stat="identity") +
  #theme(axis.text.x = element_text(angle=90)) + 
  labs(title = "", y = "Participants", x = "Active duration (%)") +
  scale_fill_discrete("", labels = c('Fairly Active', 'Light Active', 'Very Active'))
  #scale_fill_manual("", values = c('FairlyActivePerc'='blue', 'LightlyActivePerc'='red', 'VeryActivePerc'='green'), labels = c('Fairly Active', 'Light Active', 'Very Active'))
```

We also computed the averaged duration for each activity across participants and its standard errors as follows:

```{r}
activity_sum = data.frame(Purpose = c("LightlyActive","FairlyActive", "VeryActive"), Average = c(mean(activity_merged$LightlyActiveMinutesTotal), mean(activity_merged$FairlyActiveMinutesTotal), mean(activity_merged$VeryActiveMinutesTotal)), SE = c(sd(activity_merged$LightlyActiveMinutesTotal)/sqrt(length(activity_merged$ID)), sd(activity_merged$FairlyActiveMinutesTotal)/sqrt(length(activity_merged$ID)), sd(activity_merged$VeryActiveMinutesTotal)/sqrt(length(activity_merged$ID))))
activity_sum
```

The following bar plot indicates the averaged duration the participants used their devices to monitor each activity in minutes. 

```{r}
ggplot(activity_sum, aes(x=Purpose, y=Average)) + 
  geom_bar(stat="identity", position=position_dodge()) + 
  geom_errorbar(aes(ymin=Average-SE, ymax=Average+SE), width=.1, position=position_dodge(.9)) +
  labs(title="Averaged usage for tracking activity in 30 days across participants", x="Purpose", y="Averaged active duration [min]")
```

To perform a statistical test, we made another long data table that contained total duration for each activity for each individual as follows:

```{r}
active_min_wide = summarize(activity_merged, ID, LightlyActiveMinutesTotal, FairlyActiveMinutesTotal, VeryActiveMinutesTotal)
active_min_long = gather(active_min_wide, Activity, Duration, -ID)
active_min_long = mutate(active_min_long, char_ID = as.character(active_min_long$ID))
head(active_min_long)
```

As the bar plot indicated, all the participants tended to use their devices for tracking light activity. An one-way repeated measures ANOVA revealed that there was a statistically singificant main effect of the type of activity (p < 0.001) as follows:

```{r}
one.way <- aov(Duration ~ Activity, data = active_min_long)
summary(one.way)
```

### Summary

The results showed that people are nearly only interested in tracking light activity rather than moderate and intensive activities using their FitBit, which supported our hypothesis. The averaged duration for tracking light activity across participants was 5492.21 $\pm$ 405.90 min, which was about 85% of the total activity tracking duration, and those for tracking moderate and intensive activities were 386.29 $\pm$ 63.82 min and 601.88 $\pm$ 128.09 min, respectively. 

## General discussions and recommendation

In this project, we analyzed the FitBit usage data to investigate how FitBit users have been using their devices to inform Bellabeat's marketing stragety. The analysis revealed that:

* FitBit users are more interested in monitoring their active duration.
* FitBit users are in general lightly active.

Assuming that potential users of FitBit and Bellabeat's devices are greatly overlapped, this analytics project suggested that the potential users of Bellabeat's products would be more interested in monitoring casual activities (e.g., walking). Therefore, Bellabeat should put in more efforts on implementing products, functionality and services related to monitoring and enhancing activity and advertise them. For example:

* Produce many different types of devices that can be worn casually and looks nice on casual clothes. The existing necklace and bracelet devices would be a Bellabeat's strenth as FitBit does not have any devices like them. Improving the design of the devices, e.g., by collaborating with famous fashion brands, would attract more customers.

* Add functions to keep up users' motivation to be active. For example, when a device detect that an user is not active, the device reminds him/her to stand up, go for a walk, and so on.

The limitation of this analysis is that we did not have data related to the users' stress tracking. Therefore, there might be some users who are more interested in tracking their stress levels than their activities. 
